---
title: "Dashboard User Guide"
author: "Parth Tendulkar"
date: "6 November 2025"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dashboard User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Introduction

The `bushfire` interactive dashboard provides a comprehensive interface for exploring bushfire activity in southeastern Australia (1997–2018). This guide walks you through each feature of the dashboard and explains how to interpret the visualizations.

## Dashboard Overview

The dashboard consists of two main tabs accessible from the left sidebar:

1. **Overview** — Yearly trends of burned area and mean Fire Weather Index (FWI), with filters, table, and downloads  
2. **Correlation** — Yearly Pearson correlation between monthly FWI and monthly burned area, with filters, table, and downloads

Each tab is self-contained with its own controls, visualizations, and interpretation guides.

---

## Tab 1: Overview

### Purpose

The Overview tab presents yearly summaries compiled from monthly observations, showing:

- How total burned area (km²) changes over time
- How mean FWI evolves across years
- Side-by-side review using a combined bar (burned area) + line (FWI) plot

### Key Features

#### Value Boxes

Three summary statistics provide a quick overview:

- **Total Monthly Records**: Number of monthly observations in the dataset
- **Average Fire Weather Index**: Mean FWI across all months
- **Total Burned Area (km²)**: Sum of burned area across all months

These boxes give you an at-a-glance understanding of dataset size, typical conditions, and overall impact.

#### Information Box (About Data)

A collapsible information panel explains key concepts:

- **FWI (Fire Weather Index)**: Composite indicator combining wind, humidity, temperature, and rainfall. Higher values indicate conditions more conducive to fire ignition and spread.
- **Burned Area (km²)**: Total land area affected by bushfires.
- **Plot interpretation**: Bars represent annual burned area; the blue line represents mean FWI on a secondary axis.

#### Interactive Trend Visualization (Bar + Line)

The main feature is a combined plot showing yearly **total burned area** (bars, primary y-axis) and **mean FWI** (line, secondary y-axis):

**How to interact:**

- **Hover**: See exact values per year
- **Zoom/Pan**: Use plotly toolbar controls
- **Reset**: Click the home icon to return to full view

**Scaling note:**

- The line uses an internal scaling (e.g., multiplying mean FWI by 1000) so it can be shown alongside burned area while FWI remains interpretable via the **secondary axis**.

**What to look for:**

- **Peaks**: Identify years with unusually high burned area
- **Co-movement**: Check if years with higher mean FWI align with larger burned area
- **Period focus**: Use the year slider to examine specific windows (e.g., early vs later years)

#### Filter Panel

- **Year Range**: Adjust the slider to focus on years of interest.  
  Minimum and maximum values are set from the data and update automatically if the dataset changes.

#### Yearly Summary Table

A clean, paginated table lists yearly totals and means used in the plot:

- Sort columns by clicking headers
- Scan values to corroborate visual trends
- Default pagination shows 10 rows for readability

#### Download Data

Convenient CSV downloads for reproducible analysis:

- **Monthly CSV** — Full monthly dataset used by the dashboard
- **Yearly Summary CSV** — Yearly totals and means
- **Population CSV** — Reference population table included for context

Filenames include the current date for versioning and traceability.

### Interpretation Tips

1. **FWI vs activity**: Higher FWI often aligns with larger burned area, but other factors (ignitions, suppression, fuel) also matter.  
2. **Outliers**: Extremely high burned area years may reflect large events; cross-check with external records if needed.  
3. **Trend shifts**: Multi-year patterns can reflect climatic variability or operational changes (e.g., management, reporting).

---

## Tab 2: Correlation

### Purpose

This tab explores how closely monthly **FWI** and **burned area** move together **within each year**, using **Pearson correlation**:

- For each year, correlation is computed across that year’s monthly observations
- The result is a time series of yearly correlation values (r)

### Key Features

#### Summary Value Box

- **Overall Correlation (FWI vs Burned Area)**: Pearson r computed using all monthly observations across all years.  
  - **+1** strong positive, **0** none, **−1** strong negative

Use this to gauge the aggregate relationship before drilling into year-level detail.

#### Interactive Correlation Plot

A line chart with points displays **yearly r** values on the y-axis (bounded −1 to 1):

**How to interact:**

- **Hover**: View the year and corresponding correlation value
- **Zoom/Pan**: Use plotly toolbar
- **Reset**: Click home icon

**What to look for:**

1. **Consistency**: Are correlations mostly positive (or negative) across the period?  
2. **Strength**: Which years show notably strong or weak relationships?  
3. **Regime changes**: Sudden flips or trends may indicate changes in climate, fuels, or data coverage.

#### Filter Panel

- **Year Range**: A dedicated slider lets you narrow the subset of years used in the plot and correlation download.

#### Supporting Table & Download

- **Population Summary**: A companion table providing contextual population estimates (no paging).  
- **Download Correlation Data**: Export the per-year correlation series as CSV (filename includes the current date).

### Interpretation Tips

- **Correlation ≠ causation**: Alignment between weather index and burned area supports—but doesn’t prove—weather-driven dynamics.  
- **Sample size & missingness**: Years with fewer valid monthly pairs can yield more volatile r values.  
- **Outlier months**: Single extreme months can influence yearly correlation; consider inspecting raw monthly data when needed.

---

## General Dashboard Features

### Navigation

- **Sidebar menu**: Click **Overview** or **Correlation** to switch tabs
- **Active tab**: Highlighted in the sidebar for orientation
- **Collapsible boxes**: Use titles to expand/collapse information panels

### Interactivity

All visualizations are interactive:

- **Hover** for details
- **Zoom/Pan/Reset** via plotly toolbar
- **Tables**: Sort columns for quick comparisons

### Performance Tips

For the best experience:

1. **Use modern browsers**: Chrome, Firefox, Safari, or Edge  
2. **Close unused tabs**: Reduces memory usage  
3. **Narrow filters**: Limit the year range on the Correlation tab if rendering is slow on modest hardware

---

## Common Questions

### Why do some correlations vary widely year to year?

- Differences in ignition sources, suppression, fuel conditions, or climate drivers
- Smaller numbers of valid monthly pairs in some years
- Influence of extreme months

### What does the secondary axis on the Overview plot represent?

The **right-hand axis** is the original **FWI** scale. Internally, the line is scaled for visualization, but interpretation should be based on the secondary axis labels.

### Can I export the data?

Yes. Use the **Download** buttons to export:

- Monthly data (`bushfire`)
- Yearly summary (`bushfire_summary`)
- Population table (`pop_est`)
- Yearly correlations (computed by the app)

---

## Best Practices for Dashboard Use

### 1. Start Broad, Then Narrow

- Begin with **Overview** for context
- Use the **Correlation** tab to examine alignment between weather and burned area

### 2. Consider Multiple Drivers

- Weather (FWI) is important, but ignition sources, fuel loads, and suppression strategies also shape outcomes

### 3. Watch the Scales

- Burned area is shown on the **left y-axis (km²)**
- FWI appears on the **right y-axis** with a scaled line

### 4. Validate Insights

- Cross-check peaks and correlation changes against known event years or external datasets

### 5. Document & Export

- Note interesting periods
- Export CSVs for reproducibility and deeper analysis

---

## Glossary (Quick Reference)

- **FWI (Fire Weather Index)**: Composite index of fire danger conditions (wind, humidity, temperature, rainfall).  
- **Burned Area (km²)**: Total area affected by bushfires.  
- **Pearson Correlation (r)**: Measure of linear association (−1 to +1).  
- **DALYs/YLL/YLD**: *Not used in this dashboard* — terms from population health metrics (included here only for users familiar with such indicators).

---

## Session Info

```{r session-info}
sessionInfo()
```
